### 2022-03-01 iwsm/experiments.tex
Edit some more

We apply our proposed methods to a variety of labelled networks:

\begin{itemize}
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=3$) -- network of Amazon political books published close to the 2004 presidential election. Two books are connected if they were frequently co-purchased. Vertex features encode the political affiliation of the author (liberal, conservative, or neutral).
	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of 238 individuals (students and teachers),
with edges denoting face-to-face contacts at a primary school in Lyon, France. 
The vertex features are class membership (one of 10 values: 1A-5B), gender (male, female), and status (teacher, student). We choose to analyse just the second day of results.
	\item \textbf{Facebook egonet} \cite{fb-snap} ($N=747, E=30025, D=480$) -- network of Facebook users with edges denoting ``friends''.
Vertex features are fully anonymised and encode information about each user's education history, languages spoken, gender, home-town, birthday etc. We focus on the egonet with id 1912.
\end{itemize}
%
For reference, the inferred partitions for all of these are given on Figure~\ref{fig:graphs-all}.
We employ the following metrics to assess model performance. 
First, the average
description length per entity (nodes and edges) 
$\bar{S}_e$ 
used to gauge the SBM fit is defined as:
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right).
	\label{eqn:mean-dl}
\end{equation}
%
Next, to assess the performance of the feature-to-block predictor, 
the vertex set $[N]$ 
is partitioned at random so that 
a constant fraction $f$ of vertices form the training set $\Gcal_0$ and 
the remainder form the test set $\Gcal_1$.
The $b$-chain is run using the whole network but only vertices $v \in \Gcal_0$ are 
used for the $\theta$-chain. 
Then the  the average cross-entropy loss 
over each set is used to gauge the quality of the fit,
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star^{(t)},
	\quad \textrm{where} \quad
	\Lcal_\star^{(t)} \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
	\label{eqn:cross-entropy-loss}
\end{equation}
%
where $\star \in \{0, 1\}$ toggles between the training and test sets
and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%
Nevertheless, the cross-entropy loss is a coarse measure of fit. 
A new measure, specific to each detected block,
can be defined as follows. Let
$\Bcal_\star(j)$ 
be the set of vertices with maximum a posteriori probability of belonging 
to block~$j$,
$
	\Bcal_\star(j) \coloneqq \{i \in \Gcal_\star : \hat{b}_i = j\},
$
where
$ 
	\hat{b}_i \coloneqq \argmax_j \hat{y}_{ij},
$
and
define the {\em block-accuracy} for block $j$ as,
%
\begin{equation}
	\eta_\star(j) \coloneqq \frac{1}{|\Bcal_\star (j)| \cdot 
	|\Tcal_\theta| } 
	\sum_{i \in \Bcal_\star (j)}  \sum_{t \in \Tcal_\theta}
	\one \left\{\hat{b}_i = \argmax_j \phi_j \left( x_i; \theta^{(t)} \right) \right\}.
	\label{eqn:accuracy}
\end{equation}
%
This effectively tests whether the feature-to-block and 
graph-to-block predictions agree in their largest component.
For the higher-dimensional datasets, we also apply the 
dimensionality reduction method 
of Section~\ref{sec:dim-reduction}.  
We then retrain the feature-block predictor using only the retained 
feature set $\Dcal'$, and report the log-loss over the training and 
test sets for the reduced classifier -- 
denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. 
\input{experiments-figures}
\input{experiments-detailed}

>>>>

We apply our proposed methods to a variety of real-world datasets.
For reference, the inferred partitions for all of these are given on Figure~\ref{fig:graphs-all}.
%We employ the following metrics to assess model performance. 
%First, the average
%description length per entity (nodes and edges) 
%$\bar{S}_e$ 
%used to gauge the SBM fit is defined as:
%%
%\begin{equation}
%	\bar{S}_e \triangleq \frac{1}{(N+E) |\mathcal{T}_b|} \sum_{t\in \mathcal{T}_b} S \left( b^{(t)} \right).
%	\label{eqn:mean-dl}
%\end{equation}
%%
%Next, to assess the performance of the feature-to-block predictor, 
%the vertex set $[N]$ 
%is partitioned at random so that 
%a constant fraction $f$ of vertices form the training set $\mathcal{G}_0$ and 
%the remainder form the test set $\mathcal{G}_1$.
%The $b$-chain is run using the whole network but only vertices $v \in \mathcal{G}_0$ are 
%used for the $\theta$-chain. 
%Then the  the average cross-entropy loss 
%over each set is used to gauge the quality of the fit,
%%
%\begin{equation}
%	\bar{\mathcal{L}}_\star \triangleq \frac{1}{|\mathcal{T}_\theta|} \sum_{t \in \mathcal{T}_\theta} \mathcal{L}_\star^{(t)},
%	\quad \textrm{where} \quad
%	\mathcal{L}_\star^{(t)} \triangleq \frac{1}{|\mathcal{G}_\star|} \sum_{i \in \mathcal{G}_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)},
%	\label{eqn:cross-entropy-loss}
%\end{equation}
%%
%where $\star \in \{0, 1\}$ toggles between the training and test sets
%and $\hat{y}_{ij}$ is defined in~(\ref{eqn:y-hat}).
%%
%Nevertheless, the cross-entropy loss is a coarse measure of fit. 
%A new measure, specific to each detected block,
%can be defined as follows. Let
%$\mathcal{B}_\star(j)$ 
%be the set of vertices with maximum a posteriori probability of belonging 
%to block~$j$,
%$
%	\mathcal{B}_\star(j) \triangleq \{i \in \mathcal{G}_\star : \hat{b}_i = j\},
%$
%where
%$ 
%	\hat{b}_i \triangleq \underset{j}{\operatorname{argmax}}\hat{y}_{ij},
%$
%and
%define the {\em block-accuracy} for block $j$ as,
%%
%\begin{equation}
%	\eta_\star(j) \triangleq \frac{1}{|\mathcal{B}_\star (j)| \cdot 
%	|\mathcal{T}_\theta| } 
%	\sum_{i \in \mathcal{B}_\star (j)}  \sum_{t \in \mathcal{T}_\theta}
%	\boldsymbol{1} \left\{\hat{b}_i = \underset{j}{\operatorname{argmax}}\phi_j \left( x_i; \theta^{(t)} \right) \right\}.
%	\label{eqn:accuracy}
%\end{equation}
%%
%This effectively tests whether the feature-to-block and 
%graph-to-block predictions agree in their largest component.
%For the higher-dimensional datasets, we also apply the 
%dimensionality reduction method 
%of Section~\ref{sec:dim-reduction}.  
%We then retrain the feature-block predictor using only the retained 
%feature set $\mathcal{D}'$, and report the log-loss over the training and 
%test sets for the reduced classifier -- 
%denoted $\bar{\mathcal{L}}_0'$ and $\bar{\mathcal{L}}_1'$ respectively.

### 2022-03-01 iwsm/inference.tex
Edit some more

Having completed the definition of the FFBM, we wish to leverage it 
to perform inference. Specifically, given a labelled network $(A, X)$, we wish to infer if and how the observed features $X$ impact the graphical structure $A$. Formally,
this means characterising the posterior distribution:
$
p(\theta|A, X) \propto p(\theta) \cdot p(A | X, \theta).
$
Although the prior is easily computable, 
computing the likelihood 
requires summing over all latent block-states, 
$p(A| X, \theta) = \sum_{b \in [B]^N} p(A | b) P(b | X, \theta)$, which is 
clearly impractical. In fact, this
approach is doubly intractable as we would also 
need to compute the normalising constant $p(A|X)$.
Therefore, following standard Bayesian practice,
instead we aim to draw samples from the posterior,
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X).
\end{equation}
%
We propose an iterative Markov chain Monte Carlo
(MCMC) approach to obtain these samples
$\{\theta^{(t)}\}$. We first draw a sample $b^{(t)}$ 
from the block membership posterior,
and then use $b^{(t)}$ to obtain a corresponding
sample $\theta^{(t)}$:
%
\begin{equation}
	b^{(t)} \stackrel{\rm distr}{\approx} p \big( b | A, X \big) 
	\quad \textrm{then} \quad
	\theta^{(t)} \stackrel{\rm distr}{\approx} 
	p\big(\theta | X, b^{(t)} \big),
\end{equation}
%
where these approximations become exact as
the number of MCMC iterations $t\to\infty$.
As described in the following subsections,
this can be implemented through a two-level
Markov chain via the Metropolis-Hastings (MH) 
algorithm \cite{hastings-alg}.
The splitting of the Markov chain into two levels allows us to side-step the summation over
all latent $b \in [B]^N$ required to directly compute the likelihood, $p(A| X, \theta)$.
The resulting $\theta^{(t)}$ samples are asymptotically
unbiased in that the expectation of 
their distribution converges to the true posterior:
%
\begin{equation}
\lim_{t\to\infty}
\mathbb{E}_{b^{(t)}} \left[p \left( \theta | X, b^{(t)} \right) \right] = \sum_{b \in [B]^N} p(\theta | X, b) p(b | A, X) = p(\theta | A, X).
\label{eqn:theta-unbiased}
\end{equation}
%
This is an example of a pseudo-marginal approach;
see, e.g., Andrieu and Roberts~\cite{pseudo-marginal} 
for a detailed rigorous derivation based on~(\ref{eqn:theta-unbiased}).
%
\begin{figure}[!ht]
	\centering

@@

Figure \ref{fig:samp-sequence} shows an overview of the proposed method, with $q$ and $\alpha$ denoting the MH proposal distribution and acceptance probability respectively.
Note the importance of the simplification in~(\ref{eqn:b-pseudo-prior}). 
As evaluating $p(b| X)$ does not depend on $X$, 
we do not need $X$ to sample $b$.
And on the other level, in order to obtain 
samples for $\theta$
we use only $b$ but not $A$, as $(\theta \perp \!\!\! \perp A )| b$.

>>>>

Having completed the definition of the FFBM, we wish to leverage it 
to perform inference. Specifically, given a labelled network $(A, X)$, we wish to infer if and how the observed features $X$ impact the graphical structure $A$. Formally,
this means characterising the posterior distribution:
$
p(\theta|A, X) \propto p(\theta) \cdot p(A | X, \theta).
$
Although the prior is easily computable, 
computing the likelihood $p(A| X. \theta)$
requires summing over all $B^N$ latent block-states which is 
clearly impractical. In fact, this
approach is doubly intractable as we would also 
need to compute the normalising constant $p(A|X)$.
Therefore, following standard Bayesian practice,
instead we aim to draw samples from the posterior,
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(t)} \sim p(\theta | A, X).
\end{equation}
%
We propose an iterative Markov chain Monte Carlo
(MCMC) approach to obtain these samples
$\{\theta^{(t)}\}$. We first draw a sample $b^{(t)}$ 
from the block membership posterior,
and then use $b^{(t)}$ to obtain a corresponding
sample $\theta^{(t)}$:
%
\begin{equation}
	b^{(t)} \stackrel{\rm distr}{\approx} p \big( b | A, X \big) 
	\quad \textrm{then} \quad
	\theta^{(t)} \stackrel{\rm distr}{\approx} 
	p\big(\theta | X, b^{(t)} \big),
\end{equation}
%
where these approximations become exact as
the number of MCMC iterations $t\to\infty$.
As described in the following subsections,
this can be implemented through a two-level
Markov chain via the Metropolis-Hastings (MH) 
algorithm [Hastings (1970)].
The splitting of the Markov chain into two levels allows us to side-step the summation over
all latent $b \in [B]^N$ required to directly compute the likelihood, $p(A| X, \theta)$.
The resulting $\theta^{(t)}$ samples are asymptotically
unbiased in that the expectation of 
their distribution converges to the true posterior:
%
\begin{equation}
\lim_{t\to\infty}
\mathbb{E}_{b^{(t)}} \left[p \left( \theta | X, b^{(t)} \right) \right] = \sum_{b \in [B]^N} p(\theta | X, b) p(b | A, X) = p(\theta | A, X).
\label{eqn:theta-unbiased}
\end{equation}
%
This is an example of a pseudo-marginal approach;
see, e.g., Andrieu and Roberts (2009) 
for a detailed rigorous derivation based on~(\ref{eqn:theta-unbiased}).
%
\begin{figure}[!ht]
	\centering

@@

Figure \ref{fig:samp-sequence} shows an overview of the proposed method, with $q$ and $\alpha$ denoting the MH proposal distribution and acceptance probability respectively.
Due to our formulation, evaluating $p(b| X)$ does not depend on $X$ so
we do not need $X$ to sample $b$.
And on the other level, in order to obtain 
samples for $\theta$
we use only $b$ but not $A$, as $(\theta \perp \!\!\! \perp A )| b$.

### 2022-03-01 iwsm/introduction.tex
Edit some more

Many real-world networks exhibit strong community structure, with most nodes belonging to densely connected clusters. 
Finding ways to recover the latent communities from the observed graph is an important
task in many applications,
including
compression \cite{cluster-compression} and link prediction 
\cite{link-prediction}.
In this work, we examine vertex-labelled networks, 
referring to the labels as {\em features}. A typical goal is to determine whether a given feature impacts graphical structure. Answering this requires a random graph model; the standard is the stochastic block model (SBM) \cite{vanilla-sbm}. Numerous variants of the SBM  have been proposed, e.g., the MMSBM \cite{mixed-membership-sbm} and OSBM \cite{overlapping-sbm}, but these do not include features in the graph generation process.

>>>>

Many real-world networks exhibit strong community structure, with most nodes belonging to densely connected clusters. 
Finding ways to recover the latent communities from the observed graph is an important
task in many applications,
including
compression [Abbe (2016)] and link prediction 
[Zhu et al. (2016)].
In this work, we examine vertex-labelled networks, 
referring to the labels as {\em features}. A typical goal is to determine whether a given feature impacts graphical structure. Answering this requires a random graph model;
the standard is the stochastic block model (SBM) [Nowicki and Snijders (2001)]. Numerous variants of the SBM  have been proposed, e.g., the MMSBM [Airoldi et al. (2009)] and OSBM [Zhu et al. (2016)], but these do not include features in the graph generation process.

### 2022-03-01 iwsm/prelim.tex
IWSM full scaffold
Edit some more

We first need a model for community-like structure in a network. For this we adopt the widely-used stochastic block model (SBM). This is a latent variable model where each vertex belongs to a single block and the probability two vertices are connected depends only on the block memberships of each.
Specifically, we will use the microcanonical variant of the SBM, proposed by Peixoto \cite{Peixoto-Bayesian-Microcanonical}. To allow for degree-variability between members of the same block, we adopt the following degree-corrected 
formulation (DC-SBM):
% , defined in~(\ref{defn:microcan-dc-sbm}).

\begin{definition}[Microcanonical DC-SBM]
	\label{defn:microcan-dc-sbm}
	Let $N \geq 1$ denote the number of vertices in an undirected graph
with $E$ edges. The block memberships are encoded by a vector $b \in [B]^N$,
where $B$ is the number of non-empty blocks.\footnote{For each integer $K\geq 1$, we use the notation $[K]:=\{1,2,\ldots,K\}$.}
	Let $e=(e_{rs})$ be the $B \times B$ symmetric matrix of edge counts 
between blocks, such that $e_{rs}$ is the number of edges from block $r$ to 
block $s$. 
	Let $k =(k_i)$ denote a vector of length $N$, with $k_i$ being the degree of vertex $i$.

The graph's adjacency matrix $A \in \{0,1\}^{N \times N}$ is generated 
by placing edges uniformly at random, conditional
on the constraints imposed by $b$, $e$ and $k$ being satisfied.
Specifically, if $A \sim \mbox{\rm DC-SBM}_{\rm MC} (b, e, k)$,
then with probability~1 it satisfies,
for all $r,s\in[B]$
and all $i\in[N]$:
	%
	\begin{equation}
		e_{rs} = \sum_{i, j \in [N]} A_{ij} 
	\one \{b_i = r\} \one \{b_j = s\} 
		\qquad 
		\textrm{and} \qquad
		k_i = \sum_{j \in [N]} A_{ij}.
		\label{eqn:sbm-constraints}
	\end{equation}
\end{definition}

>>>>

We first need a model for community-like structure in a network. For this we adopt the widely-used stochastic block model (SBM). This is a latent variable model where each vertex belongs to a single block and the probability two vertices are connected depends only on the block memberships of each.
Specifically, we will use the microcanonical variant of the SBM, proposed by Peixoto (2016). To allow for degree-variability between members of the same block, we adopt the following degree-corrected 
formulation (DC-SBM):
% , defined in~(\ref{defn:microcan-dc-sbm}).

\textbf{Microcanonical DC-SBM} --
Let $N \geq 1$ denote the number of vertices in an undirected graph
with $E$ edges. The block memberships are encoded by a vector $b \in [B]^N$,
where $B$ is the number of non-empty blocks.\footnote{For each integer $K\geq 1$, we use the notation $[K]:=\{1,2,\ldots,K\}$.}
	Let $e=(e_{rs})$ be the $B \times B$ symmetric matrix of edge counts 
between blocks, such that $e_{rs}$ is the number of edges from block $r$ to 
block $s$. 
	Let $k =(k_i)$ denote a vector of length $N$, with $k_i$ being the degree of vertex $i$.

The graph's adjacency matrix $A \in \{0,1\}^{N \times N}$ is generated 
by placing edges uniformly at random, conditional
on the constraints imposed by $b$, $e$ and $k$ being satisfied.
Specifically, if $A \sim \mbox{\rm DC-SBM}_{\rm MC} (b, e, k)$,
then with probability~1 it satisfies,
for all $r,s\in[B]$
and all $i\in[N]$:
%
\begin{equation}
	e_{rs} = \sum_{i, j \in [N]} A_{ij} 
\boldsymbol{1} \{b_i = r\} \boldsymbol{1} \{b_j = s\} 
	\qquad 
	\textrm{and} \qquad
	k_i = \sum_{j \in [N]} A_{ij}.
	\label{eqn:sbm-constraints}
\end{equation}
